Staged Abstract Interpreters

### 2019-09-16 paper/10_intro.tex
misc

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an operational
approach to constructing abstract interpreters.  By deriving semantics artifacts
for abstract interpretation from their concrete counterparts (for example,
abstract machines), soundness can be easily established by examining the
transformation of semantic artifacts. This systematic approach to abstraction
can be tailored to different language features (such as state, first-class
control, exceptions, etc.) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
Darais:2015:GTM:2814270.2814308}. It has also been applied to various small-step
abstract machines \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
Sergey:2013:MAI:2491956.2491979} and big-step definitional interpreters
\cite{Wei:2018:RAA:3243631.3236800, DBLP:journals/pacmpl/DaraisLNH17,
Keidel:2018:CSP:3243631.3236767}.
Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows \cite{Keidel:2018:CSP:3243631.3236767},
extensible effects \cite{Githubsemantic},
etc. These pure approaches provide certain benefits, including the fact that the
abstract interpretation artifacts can be built in a compositional and modular
fashion. Furthermore, referential transparency allows programmers to confidently
reason about the correctness. The soundness of an
analysis can also be proven with less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of interpreted languages. For example, the list monad can be used to model
nondeterminism of collecting interpreters. Although such pure approaches have
their merits, they are significantly slower when compared with imperative,
stateful implementations. The goal of this paper is to develop an approach for
implementing abstract interpreters, which allows high-level abstractions during
implementation for clarity, but passes the incurred overhead at runtime for
efficiency.
%In order to reconcile these approaches, one need to
%use a programming technique which allows for high-level abstractions during
%implementation, but bypasses the incurred overhead at runtime.

>>>>

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an operational
approach to constructing abstract interpreters.  By deriving semantics artifacts
for abstract interpretation from their concrete counterparts (for example,
abstract machines), soundness can be easily established by examining the
transformation of semantic artifacts. This systematic approach to abstraction
can be tailored to different language features (such as state, first-class
control, exceptions, etc.) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
Darais:2015:GTM:2814270.2814308}. It has also been applied to various small-step
abstract machines \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
Sergey:2013:MAI:2491956.2491979} and big-step definitional interpreters
\cite{Wei:2018:RAA:3243631.3236800, DBLP:journals/pacmpl/DaraisLNH17,
Keidel:2018:CSP:3243631.3236767}.
Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows \cite{Keidel:2018:CSP:3243631.3236767},
and extensible effects \cite{Githubsemantic}.
These pure approaches provide certain benefits, including the fact that the
abstract interpretation artifacts can be built in a compositional and modular
fashion. Furthermore, referential transparency allows programmers to confidently
reason about the correctness. The soundness of an
analysis can also be proven with less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of interpreted languages. For example, the list monad can be used to model
nondeterminism of collecting interpreters. Although such pure approaches have
their merits, they are significantly slower when compared with imperative,
stateful implementations. The goal of this paper is to develop an approach for
implementing abstract interpreters, which allows high-level abstractions during
implementation for clarity, but bypasses the incurred overhead at runtime for
efficiency.
%In order to reconcile these approaches, one need to
%use a programming technique which allows for high-level abstractions during
%implementation, but bypasses the incurred overhead at runtime.

### 2019-09-16 paper/abstract.tex
misc

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstractions and
monadic abstractions. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in one of four modes: as an unstaged
concrete interpreter, a staged concrete interpreter, an unstaged abstract
interpreter, or a staged abstract interpreter. As an example of
\emph{abstraction without regret}, the overhead of these abstraction layers is
eliminated in the generated code after staging. We show that staging abstract
interpreters is practical and useful to optimize static analysis, all while
requiring less engineering effort and without compromising soundness. We conduct
three case studies, including a comparison with Boucher and Feeley's abstract
compilation, applications to various control-flow analyses, and a demonstration
that can be used for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.

>>>>

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
generic interpreter with type-level binding-time abstractions and
monadic abstractions. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in one of four modes: as an unstaged
concrete interpreter, a staged concrete interpreter, an unstaged abstract
interpreter, or a staged abstract interpreter. As an example of
\emph{abstraction without regret}, the overhead of these abstraction layers is
eliminated in the generated code after staging. We show that staging abstract
interpreters is practical and useful to optimize static analysis, all while
requiring less engineering effort and without compromising soundness. We conduct
three case studies, including a comparison with Boucher and Feeley's abstract
compilation, applications to various control-flow analyses, and a demonstration
for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.
